---
title: "Introduction to data preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to data preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most camera-trap analyses in `ct` share the same first steps: read the raw
image records, inspect them, and reduce the stream of photographs to a set of
*independent* detection events before any activity, diversity, or density
method is applied. This vignette walks through that preparation using the
bundled `penessoulou` dataset, a year of camera-trap records from the
Penessoulou Classified Forest in Benin.

```{r setup}
library(ct)
library(dplyr)
```

## The raw records

`penessoulou` holds one row per recorded image, with the species seen, the
camera that took it, and the capture date-time (as a `"YYYY-MM-DD HH:MM:SS"`
string).

```{r data}
data(penessoulou)

dim(penessoulou)

penessoulou %>%
  select(project, camera, species, number, datetimes) %>%
  head()
```

A quick tally shows how the raw images are distributed across species. At this
stage every photograph counts, so a single animal lingering in front of a
camera can inflate these numbers.

```{r counts}
penessoulou %>%
  count(species, sort = TRUE)
```

## Filtering independent detections

Consecutive photographs of the same species at the same camera within a short
time window are not statistically independent (Ridout & Linkie, 2009). The
`ct_independence()` function collapses such bursts, keeping detections only when
the gap since the previous record of that species at that camera meets a chosen
`threshold` (in seconds). Here we use a 30-minute (1800 s) window, assessed
separately per species and per camera.

```{r independence}
independent <- penessoulou %>%
  ct_independence(
    datetime = datetimes,
    format = "%Y-%m-%d %H:%M:%S",
    species_column = "species",
    site_column = "camera",
    threshold = 1800
  )

nrow(penessoulou)
nrow(independent)
```

Filtering removes the redundant burst photographs, leaving fewer, independent
events. The same species tally now reflects independent detections rather than
raw image counts:

```{r counts-after}
independent %>%
  count(species, sort = TRUE)
```

## Where to go next

The independent detections produced here are the common input to the rest of
the package, for example:

- `ct_fit_activity()` and `ct_overlap_estimates()` for activity patterns and
  temporal overlap between species,
- `ct_camera_day()` together with `ct_alpha_diversity()` or `ct_inext()` for
  diversity and rarefaction,
- the density estimators (`ct_fit_ds()`, `ct_fit_rest()`, `ct_fit_tte()`,
  `ct_fit_ste()`).

Each of these is covered in its own article on the
[package website](https://stangandaho.github.io/ct/).

## References

Ridout, M. S., & Linkie, M. (2009). Estimating overlap of daily activity
patterns from camera trap data. *Journal of Agricultural, Biological, and
Environmental Statistics*, 14(3), 322-337.
