---
title: "Camera Trap Data Processing"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

## Why preparation comes first

A camera trap does not hand you a tidy table. It hands you a long stream of
images, and every image becomes a row: the same animal standing in front of the
lens for a minute can leave a dozen near identical records. Before any activity
analysis, diversity index, or density model, that raw stream has to be turned
into something the method can trust. This article walks through a short preparation workflow with `ct`, from a first look at the records to a clean
site by species table ready for analysis.

We use the bundled `penessoulou` dataset, from the
Penessoulou Classified Forest in Benin. It holds one row per recorded image.

```{r setup, message=FALSE, warning=FALSE}
library(ct)
library(dplyr)

data(penessoulou)

dim(penessoulou)

penessoulou %>%
  select(project, camera, species, number, datetimes) %>%
  head()
```

Each row carries the survey phase (`project`), the camera that took the picture
(`camera`), the identified `species`, how many individuals were counted
(`number`), and the capture time as a `"YYYY-MM-DD HH:MM:SS"` string
(`datetimes`), along with the camera coordinates. This dataset is direct output of 
[Declas](https://github.com/stangandaho/declas), a species detection and classification software designed to handle efficiently camera trap data.

## A first look at the data

Before touching anything, it helps to know what you are holding. `ct_describe_df()`
summarises whichever columns you name, giving frequencies and proportions for
categorical variables and the usual descriptive statistics for numeric ones.

```{r describe}
ct_describe_df(penessoulou, species, number)
```

By default `ct_describe_df()` summarises the numeric column within each level of
the categorical one. Here `number` is the count of individuals visible on an
image, so this single call lists every recorded species, how many images hold it
(`N`), and how many individuals typically appear per image (the `Min`, `Max`,
`Median` and `Mean` of `number`). You can also pass your own function, for
example the sum, to add the sum number of individuals per species. If you
would rather see plain frequency tables and one overall numeric summary, pass
`by_group = FALSE`.

## Keeping only independent detections

Two photographs of the same species at the same camera taken seconds apart are
not two pieces of evidence. They are one event seen twice. Counting them
separately inflates activity and abundance and breaks the independence that most
methods assume. `ct_independence()` collapses these bursts: it keeps a detection
only when enough time has passed since the previous record of that species at
that camera. Here we use a 30 min window (1800 seconds), applied
separately within each species and each camera.

```{r independence}
independent <- penessoulou %>%
  ct_independence(
    datetime = datetimes,
    format = "%Y-%m-%d %H:%M:%S",
    species_column = "species",
    site_column = "camera",
    threshold = 1800
  )

nrow(penessoulou)
nrow(independent)
```

The stream of images collapses to a much smaller set of independent events. The
threshold is a choice, not a constant: thirty minutes is common (e.g. Ahmad et al., 2024), but a study on a fast moving species or a busy trail might use a shorter or longer window. 
The point is to make that choice deliberately and to write it down.

## Checking how the cameras ran

Independent or not, a detection is only meaningful if the camera was actually
working. A camera that stops for two weeks tells a very different story from one
that ran without a break, yet both can look similar in a raw table.
`ct_plot_camtrap_activity()` draws each camera's active window along time and
marks any gap longer than a threshold you set, so you can see the sampling
effort at a glance.

```{r activity, message=FALSE, warning=FALSE, fig.height=6, fig.alt="Timeline showing the active period of each camera during the survey, with one horizontal bar per camera."}
penessoulou %>%
  filter(project == "Last") %>%
  ct_plot_camtrap_activity(
    deployment_column = camera,
    datetime_column = datetimes,
    threshold = 7,
    time_unit = "days",
    ybreak = "2 days"
  )
```

Each bar is the span between a camera's first and last record for this survey
phase. Gaps beyond the threshold, had there been any, would appear as red
segments. A picture like this is a quick data quality check: it exposes cameras
that failed early, started late, or went silent in the middle of the survey,
before those problems quietly bias an estimate.

## Reading the gaps as a table

The plot is good for spotting trouble at a glance, but sometimes you need the
exact figures. `ct_find_break()` returns every interruption in a datetime series
as a table of `start`, `end` and `duration`, so you can say precisely when a
camera fell silent and for how long. Run it one camera at a time, since a gap
only means something within a single device.

```{r find-break, warning=FALSE}
penessoulou %>%
  filter(project == "Last", camera == "CAMERA 5") %>%
  ct_find_break(
    datetime_column = datetimes,
    threshold = 2,
    time_unit = "days"
  )
```

Every stretch of more than two days without a picture is listed with its length
in days. Lower the `threshold` to surface short interruptions, raise it to keep
only the serious outages. This is the same gap detection that draws the red
segments in the plot above.

## Building a site by species table

Many downstream tools, from diversity indices to ordination, expect a community
matrix: one row per site, one column per species, and an abundance in each cell.
`ct_to_community()` reshapes the independent detections into exactly that. Because
`number` counts the individuals on each image, summing it per camera and species
gives the total individuals recorded, which we use as the abundance. Empty
combinations are filled with zero.

```{r community}
community <- independent %>%
  ct_to_community(
    site_column = camera,
    species_column = species,
    size_column = number,
    values_fill = 0
  )

dim(community)

community[1:5, 1:5]
```

Now every camera is a row and every species a column, with a real zero wherever
a species was never detected.

## Putting sites on a comparable scale

Raw counts are hard to compare across cameras, because a camera on a busy trail
records more of everything than a camera in a quiet corner. Standardisation
removes that effect so differences reflect community composition rather than raw
activity. `ct_standardize()` offers many methods; the Hellinger transformation is
one of choices before distance based analyses. It works on a numeric matrix, so
we move the camera names into row names first.

```{r standardize}
mat <- community %>%
  tibble::column_to_rownames("camera")

community_hel <- ct_standardize(mat, method = "hellinger")

round(as.matrix(community_hel)[1:5, 1:5], 3)
```

The values are now on a common footing, ready for a dissimilarity matrix, a
clustering, or an ordination without one crowded camera dominating the result.

## Detection histories for occupancy

Occupancy models ask a different question: not how many animals, but whether a
species was seen at all in each sampling window. `ct_to_occupancy()` reshapes the
detections into that format, splitting the study period into fixed windows and
recording presence (1) or absence (0) for every site and species. We take the day
straight from the parsed `datetime` column that `ct_independence()` added, which
is cleaner than the raw `dates` field, then group the survey into monthly windows.

```{r occupancy}
occupancy <- independent %>%
  mutate(day = as.Date(datetime)) %>%
  ct_to_occupancy(
    date_column = day,
    site_column = camera,
    species_column = species,
    size_column = number,
    by_day = 30,
    presence_absence = TRUE
  )

occupancy[1:5, 1:5]
```

Each row is a site and species, each window column holds a 1 or a 0, and this is
exactly the detection history an occupancy model expects. Set
`presence_absence = FALSE` to keep the raw counts instead.

## When the clock is wrong

Cameras drift. A unit set with the wrong date, or one whose battery reset the
clock, stamps every image with a time that never happened, and that quietly
poisons independence filtering, activity patterns and effort. `ct_correct_datetime()`
repairs this from a small correction table that says, for each camera, whether it
was running fast or slow and what the true time was at a reference moment.

```{r correct-datetime}
# One row per camera to fix: the direction (+ slow, - fast) and a reference
# timestamp showing the correct time.
corrector <- data.frame(
  camera = c("CAMERA 1", "CAMERA 2"),
  sign = c("+", "-"),
  datetimes = c("2025-03-14 8:17:00", "2024-11-14 10:02:03")
)

penessoulou %>%
  filter(project == "Last") %>%
  ct_correct_datetime(
    datetime = datetimes,
    deployment = camera,
    corrector = corrector
  ) %>%
  select(camera, datetimes, corrected_datetime, time_offset_seconds) %>%
  slice_head(n = 5)
```

The function adds a `corrected_datetime` column, along with the offset it applied,
while leaving the original untouched, so you can inspect the shift before trusting
it. In a real workflow this comes first, and the corrected column feeds every step
above.

## Effort and trap rate

Counts mean little without knowing how long each camera watched. Effort is the
number of camera days behind the detections, and it is the denominator for trap
rate, occupancy and most density models. `penessoulou` carries no deployment
table, so here we switch to the bundled Camtrap DP example (`ctdp`), which records
when each camera was switched on and off.

```{r effort, message=FALSE, warning=FALSE}
data("ctdp")
deployments <- ctdp$data$deployments

ct_get_effort(
  deployment_data = deployments,
  deployment_column = deploymentID,
  start_column = start,
  end_column = end
) %>%
  head()
```

`ct_get_effort()` returns the active duration of each deployment.
`ct_traprate_data()` goes a step further, joining those efforts to the detections
so every camera carries both a count (`n`) and its effort, the pair that
`ct_traprate_estimate()` needs.

```{r traprate, message=FALSE, warning=FALSE}
observations <- ctdp$data$observations %>%
  filter(scientificName == "Vulpes vulpes")

ct_traprate_data(
  observation_data = observations,
  deployment_data = deployments,
  use_deployment = TRUE,
  deployment_column = deploymentID,
  datetime_column = timestamp,
  start_column = start,
  end_column = end
) %>%
  head()
```

## Time of day on a circle

Activity analyses treat the time of day as an angle, because 23:59 and 00:01 are
two minutes apart, not twenty-four hours. `ct_to_radian()` maps a clock time onto
that circle, where a full day is `2 * pi`.

```{r radian}
with_radian <- independent %>%
  ct_to_radian(times = times, format = "%H:%M:%S")

with_radian %>%
  select(species, times, time_radian) %>%
  head()
```

The `time_radian` column is what activity tools such as `ct_fit_activity()` and
`ct_overlap_estimates()` consume. The reverse trip is `ct_to_time()`, handy for
reading an angle back as a clock time:

```{r to-time}
ct_to_time(c(0, pi / 2, pi, 3 * pi / 2))
```

## Where this leads

In a handful of steps the raw image stream became something dependable: explored,
filtered to independent events, checked for sampling gaps, and reshaped into a
standardised community table. From here the usual `ct` analyses follow naturally,
for example:

- `ct_fit_activity()` and `ct_overlap_estimates()` for daily activity and
  temporal overlap between species, fed by the radian times prepared above,
- `ct_alpha_diversity()` and `ct_dissimilarity()` on the community table for
  diversity and between site comparison,
- `ct_traprate_estimate()` for trap rate, and the density estimators
  (`ct_fit_ds()`, `ct_fit_rest()`, `ct_fit_tte()`, `ct_fit_ste()`) once effort
  and detections are in place.

Good preparation is not glamorous, but it is where a trustworthy result is
either won or lost. Spend the time here, and everything built on top of it stands
on solid ground.

## Reference
Ahmad, F., Mori, T., Rehan, M., Bosso, L., & Kabir, M. (2024). Applying a Random Encounter Model to Estimate the Asiatic Black Bear (Ursus thibetanus) Density from Camera Traps in the Hindu Raj Mountains, Pakistan. Biology 13, 341.


