---
title: "Directed encounter responses between species"
subtitle: "Does one species' passage change another's short-term detection at the same camera?"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
css: styles.css
vignette: >
  %\VignetteIndexEntry{Directed encounter responses between species}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 150
)
```

```{=html}
<style>
.caption {
  text-align: center;
  color: #717171;
  font-style: italic;
  font-size: 0.85em;
}
.readbox {
  border: 1px solid #cfe0d9;
  border-left: 4px solid #2A6A5A;
  background: #f2f8f5;
  border-radius: 0 8px 8px 0;
  padding: 0.8em 1.1em;
  margin: 1.2em 0;
  font-size: 0.94em;
}
.readbox p:first-child { margin-top: 0; }
.readbox p:last-child { margin-bottom: 0; }
.readbox .rb-title {
  font-weight: 700;
  color: #2A6A5A;
  letter-spacing: 0.02em;
}
@media (prefers-color-scheme: dark) {
  .readbox { background: #12241d; border-color: #2f4a40; }
}
</style>
```

## Beyond "do they overlap?"

A common question in camera trap ecology is whether two species are
active at the same times of day. An activity overlap answers that: it
compares two daily rhythms. But it is silent about *order* and *place*.
Two species can share an identical dusk peak and never once meet at the
same spot.

`ct_fit_encounter_response()` asks a different, more pointed question.
Along the timeline of a single camera:

> After one species (the **trigger**) passes a camera, does a second
> species (the **response**) become more, or less, likely to be detected
> there in the following hours, once we account for that camera's daily
> rhythm, its season, its location, and how long it was switched on?

This is a *directed* (trigger then response), *local* (same camera),
*short-term* (hours) association. It is a description of co-occurrence
in time, not a proof of interaction. We return to that distinction at
the end.

The mechanics is that each camera's uptime is sliced into short equal
slots (e.g 15min); the response detections in each slot are counted;
every slot is labelled by how long ago the trigger last passed; and the
response rate in the slots just after a trigger is compared with the
rate in "baseline" slots (long after a trigger, or before any), holding
time-of-day, month, camera, and slot length equal. The comparison
returns a **rate ratio** for each lag window: 1 means no change, above 1
means detected more than expected, below 1 less.

## Use case

We use the bundled `lama` dataset: a camera-trap survey of the Lama
Classified Forest, a semi-deciduous forest remnant in the Dahomey Gap of
southern Benin. Twenty-three cameras ran from June to December 2024 and
recorded eighteen mammal taxa. It ships as a list of two tables that
share a `camera` column.

```{r data}
library(ct)
library(ggplot2)

data(lama)

# One row per image: who, where, when (plus distance-sampling fields)
head(lama$observation[, c("camera", "species", "datetime", "distance")])

# One row per camera: when it was installed and pulled out
head(lama$deployment[, c("camera", "start", "end")])
```

The two most-recorded carnivores make a good test case. The **pardine
genet** (*Genetta pardina/maculata*) is a solitary, largely nocturnal
mesocarnivore; the **flat-headed cusimanse** (*Crossarchus
platycephalus*) is a diurnal, group-living mongoose.

```{r counts}
sort(table(lama$observation$species), decreasing = TRUE)[1:6]
```

We ask: **does a genet's passage shift the short-term detection of
cusimanse at the same camera?**

## Fitting an encounter-response model

Because the deployments span roughly six months, a fine 15-minute grid
would produce hundreds of thousands of slots. We use two-hour slots and
lag windows of 0 to 6, 6 to 12, and 12 to 24 hours, which matches the
day-scale question and keeps the model quick.

```{r fit, eval=FALSE}
fit <- ct_fit_encounter_response(
  data = lama$observation,
  deployment = lama$deployment,
  trigger = "Genetta pardina/maculata",
  response = "Crossarchus platycephalus",
  species_column = species,
  cam_column = camera,
  datetime_column = datetime,
  start_column = start,
  end_column = end,
  interval = 2 * 3600,
  lag_breaks = c(0, 6 * 3600, 12 * 3600, 24 * 3600),
  engine = "glm",
  n_boot = 0
)

summary(fit)
```

```{r fit-load, echo=FALSE}
# The model is precomputed once and shipped with the article, so the site
# builds quickly; the call above reproduces this object.
fit <- readRDS("encounter_response_fit.rds")
summary(fit)
```

We chose `engine = "glm"` deliberately. These counts are
*underdispersed* (the reported dispersion is below one, because most
two-hour slots hold zero or one cusimanse). The quasi-Poisson GLM
handles that natively; the mixed-model engine's negative-binomial
family, built for *over*-dispersion, is a poor fit here.

## Reading the result

```{r estimates}
fit$estimates[, c("lag", "rate_ratio", "lower", "upper", "p_value", "n_response_events")]
```

```{=html}
<div class="readbox">
<p><span class="rb-title">How to read this table</span></p>
<p><b>Start at the baseline.</b> The <code>No recent trigger</code> row is the yardstick: its
<code>rate_ratio</code> is always 1. Every other row is measured against it.</p>
<p><b>Read each ratio as a multiplier.</b> <code>rate_ratio</code> is how many times the baseline
detection rate the response species was seen at in that window, after the model has held time of day,
month, camera, and slot length equal. Above 1 means detected more often than expected; below 1 means
less; 1 means no change.</p>
<p><b>Trust the interval before the point.</b> If <code>lower</code> to <code>upper</code> straddles 1,
the data cannot separate that window from "no effect". If the whole interval sits above 1 (or below 1),
the signal is statistically supported. Read <code>p_value</code> as a rough flag (small = unlikely to be
a fluke), and glance at <code>n_response_events</code>: a big ratio built on a handful of detections is
fragile.</p>
<p><b>So, this table:</b> the <b>6 to 12 h</b> window has a rate ratio near 1.7 with an interval entirely
above 1, so cusimanse were detected about 1.7 times more often than expected in that window. The 0 to 6 h
and 12 to 24 h windows have intervals that cross 1, so they show no clear association.</p>
</div>
```

In the **6 to 12 hour** window after a genet passes, cusimanse are
detected roughly **1.7 times** as often as the camera's own time-of-day
and season would predict, and its confidence interval clears one (p
around 0.03). The 0 to 6 hour and 12 to 24 hour windows sit near one,
with intervals spanning it: no clear signal there.

A forest plot makes the pattern easy to read against the reference line
at one:

```{r forest, fig.cap="Rate ratios (points) with 95% confidence intervals for each lag window after a genet detection. Values above the dashed line at 1 mean cusimanse were detected more often than expected."}
est <- fit$estimates[fit$estimates$lag != "No recent trigger", ]
est$lag <- factor(est$lag, levels = rev(est$lag))

ggplot(est, aes(rate_ratio, lag)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.15,
                 colour = "#2A6A5A", linewidth = 0.8) +
  geom_point(size = 3.2, colour = "#2A6A5A") +
  scale_x_continuous("Rate ratio (log scale)", trans = "log10") +
  labs(y = NULL,
       title = "Cusimanse detection after a genet passes the same camera") +
  theme_minimal(base_size = 12)
```

At these cameras, a genet's passage is followed by a modest but
statistically supported rise in cusimanse detection within the following
half day, over and above what each camera's ordinary daily rhythm
predicts. It is the kind of delayed, same-site association that a diel
overlap coefficient cannot see.

## How robust is it?

An interval-based approximation should not hinge on the slot width you
happened to pick. Passing `interval_sensitivity` refits the rate ratios
across other widths so you can judge that directly. (The call above can
take `interval_sensitivity = c(3 * 3600, 4 * 3600)`; the fitted object
then carries a `$sensitivity` table.)

```{r sensitivity, eval=FALSE}
fit_s <- ct_fit_encounter_response(
  data = lama$observation, deployment = lama$deployment,
  trigger = "Genetta pardina/maculata", response = "Crossarchus platycephalus",
  species_column = species, cam_column = camera, datetime_column = datetime,
  start_column = start, end_column = end,
  interval = 2 * 3600, lag_breaks = c(0, 6 * 3600, 12 * 3600, 24 * 3600),
  engine = "glm", n_boot = 0, interval_sensitivity = c(3 * 3600, 4 * 3600)
)

fit_s$sensitivity
```

```{r sensitivity-load, echo=FALSE}
readRDS("encounter_response_fit.rds")$sensitivity
```

This is worth reading carefully, because it shows both what is robust
and what is not. Across two-, three-, and four-hour slots, every
early-window estimate stays above one: the *direction* of the effect,
more cusimanse than expected in the hours after a genet, is stable. But
*which* window carries the significant signal slides as the slots
coarsen, from the 6 to 12 hour window at two-hour slots toward the 0 to
6 hour window at four-hour slots. The honest conclusion is therefore the
broad one, an elevation within roughly the first half day, rather than a
claim pinned to one exact lag. Trusting the general pattern while
distrusting the precise window is exactly the discipline a discretised
approximation calls for.

Two engine notes for your own analyses:

- For smaller studies you can make the inference a **camera-block
  bootstrap** with `n_boot = 199` (the default for `engine = "glm"`). It
  resamples whole cameras, so it respects the fact that repeated slots
  at one camera are not independent. We skipped it here only because the
  six-month grid makes many refits slow.
- When counts are *over*-dispersed, `engine = "glmm"` fits camera as a
  random effect (via **glmmTMB** or **lme4**) and needs no bootstrap.

```{r glmm-example, eval=FALSE}
# Random-effects alternative (for over-dispersed data / many sparse cameras)
ct_fit_encounter_response(
  data = lama$observation, deployment = lama$deployment,
  trigger = "Genetta pardina/maculata", response = "Crossarchus platycephalus",
  species_column = species, cam_column = camera, datetime_column = datetime,
  start_column = start, end_column = end,
  interval = 2 * 3600, lag_breaks = c(0, 6 * 3600, 12 * 3600, 24 * 3600),
  engine = "glmm"
)
```

## What the model does, and does not, tell you

The result is an **association, conditional on the fitted baseline**,
never a proof of cause. Keep three limits in view.

- **Shared drivers.** The model removes steady, daily, and seasonal
  patterns, but not an *episodic* cause that draws both species
  independently: a fruiting tree, a waterhole, a spell of weather. Such
  a driver can manufacture an apparent response with no interaction at
  all. The genet-to-cusimanse signal here is consistent with shared use
  of the same productive microsites as much as with any direct
  following.
- **Detection is not presence.** A camera records a thinned, imperfect
  sample of what passes. A shift in detection rate need not be a shift
  in true site use.
- **Empty windows say nothing.** If the response never appears in a lag
  window, the function returns a blank for it rather than a fabricated
  number. A table full of blanks means the two species simply do not
  co-occur on this timescale, which is a finding in itself, not a result
  to over-read.

Used with those caveats, `ct_fit_encounter_response()` turns a pile of
timestamped detections into a compact, directed, effort-corrected
summary of how two species' visits line up in time, and a fair way to
ask whether one appears to react to the other.

## Data and citation

The `lama` dataset accompanies a multispecies camera-trap census of the
Lama Classified Forest:

> Adounké, G.R.M., Lecompte, E., Gandaho, S.M., Toyi, M.S., Azihou,
> A.F., Hugueny, B., Sinsin, B.A., Gaubert, P., & Djagoun, C.A.M.S.
> (submitted). Camera-trap distance sampling reveals density patterns
> and anthropogenic drivers of terrestrial mammals in a remnant forest
> refuge in West Africa. *Ecology and Evolution*.

```{r cite}
citation("ct")
```
