---
title: "Estimating Density With REST and RAD-REST Models"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
css: styles.css
vignette: >
  %\VignetteIndexEntry{Estimating Density Without Individual Recognition: the REST and RAD-REST Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
.caption {
  text-align: justify;
  color: #717171;
  font-style: italic;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)

options(rmarkdown.html_vignette.check_title = FALSE)
```


Density, the number of animals per unit area, is one of the most requested
quantities in wildlife ecology: it underpins red-list assessments, harvest
quotas, trend monitoring and impact evaluation. Yet for the great majority of
species it is genuinely hard to measure. The gold-standard methods (spatial
capture–recapture and its relatives) need animals to be **individually
identifiable** from natural marks or tags. Most mammals carry no such marks, so those methods simply do not apply.

Camera traps offer a way out. The challenge is that a pile of detections is not a density: the number of times a camera fires depends on how many animals are present, how long each lingers in front of the lens, how active the species is, and how big
the monitored patch is. The **Random Encounter and Staying Time (REST)** model
of [Nakashima, Fukasawa & Samejima (2018)](https://doi.org/10.1111/1365-2664.13059)
ties those pieces together into a single, estimable equation, and its
**RAD-REST** extension ([Nakashima et al. 2026](https://doi.org/10.1111/2041-210X.13059))
makes the data far cheaper to collect.

This article explains how REST and RAD-REST work, and shows the complete
workflow in the `ct` package, from raw detections to a posterior density
estimate using a small bundled example dataset.

# The theory of REST

## The focal area

The observer defines a
small **focal area** of known size $S$ directly in front of each camera (for
example a 2–3 m² patch marked during deployment). The only question asked of
each video is: **did the animal pass through that focal area, and for how
long?** Everything outside the focal area is ignored.The focal
area is fixed and measured, so the geometry is known in advance.

## The density equation

Imagine animals, at a density $D$, moving across the landscape. At any instant,
the expected number sitting inside a focal area of size $S$ is proportional to
$D \times S$, scaled by the fraction of the day the species is actually moving,
$p_{\text{act}}$. Each visiting animal stays for an average time $\bar{t}$
before leaving, so over a survey of duration $T$ the camera records

$$
\mathbb{E}[Y] \;=\; \frac{D \times S \times T \times p_{\text{act}}}{\bar{t}},
$$

where $Y$ is the number of **passes** through the focal area. Re-arranging gives
the REST density estimator:

$$
\boxed{\;\hat{D} \;=\; \frac{Y \times \bar{t}}{S \times T \times p_{\text{act}}}\;}
$$

Read literally: density rises with more passes ($Y$) and longer stays
($\bar{t}$, because a long-staying animal produces *few* entries, so a given
number of observed passes implies more animals), and falls as the focal area,
survey effort, or activity level grow. REST estimates **every term on the
right** jointly in a Bayesian model, so the uncertainty in staying time and
activity flows through into the density estimate rather than being ignored.

To do that, three sub-models are fitted together.

## 1. Staying time (a survival model)

The mean staying time $\bar{t}$ is estimated from the time each animal spends in
the focal area. There is a subtlety: when an animal is still inside the focal
area at the moment the video clip ends, its true staying time is **unknown**. We
only know it exceeds the clip length. These observations are [*right-censored*](https://en.wikipedia.org/wiki/Censoring_(statistics)),
exactly as in survival analysis.

REST therefore models staying time with a parametric survival distribution
(`lognormal`, `gamma`, `weibull` or `exponential`), letting censored records
contribute through the survival function rather than being discarded or
truncated. The choice of distribution matters and can be made objectively by
WAIC (see [`ct_rest_select_stay()`](#choosing-the-staying-time-distribution)).

## 2. Activity proportion

A camera only detects an animal that is up and moving; a sleeping animal in the
focal area is invisible to the encounter process. The **activity proportion**
$p_{\text{act}}$, the fraction of the 24-hour cycle the species is active, rescales
the estimate accordingly. `ct` offers two ways to estimate it:

- `kernel`: the fixed circular-kernel estimator of
  [Rowcliffe et al. (2014)](https://doi.org/10.1111/2041-210X.12278); fast and
  the default.
- `mixture`: a non-parametric Bayesian von Mises mixture
  ([Nakashima et al. 2026](https://doi.org/10.1111/2041-210X.13059)); slower,
  but it propagates activity uncertainty into the density posterior.

## 3. Encounters

Finally, the passes $Y$ themselves are modelled. In the **original REST** this 
the number of passes per station, modelled as negative-binomial to
absorb overdispersion.

Counting the total number of passes $Y$ is the most tedious and error-prone
part of REST labelling: a single clip may show an animal cross the focal area
several times, and analysts miscount. **RAD-REST** sidesteps this. Instead of a
total, each video is placed in a **pass-count class**:

- `y_0` = videos showing **0** passes through the focal area,
- `y_1` = videos showing **exactly 1** pass,
- `y_2` = videos showing **2** passes, and so on,

with `N` = total videos. The distribution of these classes is modelled with a
Dirichlet-multinomial (so miscounting is absorbed statistically), from which the
**mean number of passes per video** is estimated. The total passes are then
recovered internally as $Y = N \times \overline{\text{passes}}$ and fed into the
same density equation. The payoff is much lighter labelling. You only judge
"how many passes did *this* clip show?" — at the cost of needing at least three
count classes for the model to be identifiable.

## Assumptions

REST and RAD-REST rest on a few key assumptions:

- the **focal area is correctly delineated and measured** at every camera;
- animals **enter the focal area independently** and are not attracted to or
  repelled by the camera;
- the **population is closed** over the survey window; and
- staying time, activity and encounters are representative of the population
  (e.g. cameras are not deliberately placed where animals linger).

# The `ct` package

The reference implementation (the [ctrest](https://github.com/YoshihiroNakashima/ctrest) package) exposed the method as a set
of lower-level building blocks. `ct` wraps the whole workflow into a small,
consistent set of functions:

| Function | Role |
|---|---|
| `ct_rest_stay()` | Build the staying-time / censoring table |
| `ct_rest_passes()` | Aggregate per-station passes (`REST`) or pass-count classes (`RAD-REST`) |
| `ct_rest_effort()` | Add camera-trapping effort (days) per station |
| `ct_rest_activity()` | Turn detection times into radians for activity modelling |
| `ct_rest_select_stay()` | Compare staying-time distributions by WAIC |
| `ct_fit_rest()` | Fit the REST / RAD-REST density model (MCMC) |

**What's improved** is that every function checks its inputs up front and fails with a
clear, actionable message (powered by the `cli` package) instead of an obscure
error deep in the sampler. Arguments are named for readability (`stay_distribution`, `focal_area`, `density_formula`), the data flows through tidy data frames, and `ct_fit_rest()` upports covariates and random effects on staying time and density, both REST and RAD-REST, and both activity estimators — all from one call.

# A worked example

We use the small simulated dataset bundled with `ct`. `rest_detection` holds one
row per video; `rest_station` holds one row per camera with a `Habitat`
covariate.

```{r data, message=FALSE, warning=FALSE, results='hide'}
library(ct)
library(dplyr)
library(ggplot2)

data(rest_detection)
data(rest_station)

rest_detection %>% head(8)
```

```{r display_detection, echo=FALSE}
rest_detection %>% head(8) %>% kableExtra::kable("html") %>%
  kableExtra::kable_styling("striped")
```


## Step 1: Staying time

`ct_rest_stay()` pulls the staying-time and censoring columns and attaches the station covariates. `Cens = 1` flags a right-censored stay. The target species is the `Red duiker`.

```{r stay, results='hide'}
stay <- ct_rest_stay(
  detection_data = rest_detection,
  station_data = rest_station,
  stay_column = Stay, # column in rest_detection
  censor_column = Cens # column in rest_detection
)
stay %>% filter(Species == "Red duiker") %>% head(6)
```

```{r display_stay, echo=FALSE}
stay %>% filter(Species == "Red duiker") %>% head(6) %>% 
  kableExtra::kable("html") %>%
  kableExtra::kable_styling("striped")
```

The plot below shows that the longest stays are disproportionately censored (the animal
was still present when the clip ended), exactly the records a naive mean would
bias downward.

```{r stay-plot, fig.cap="Figure 1: Distribution of staying times for Red duiker."}
stay %>%
  filter(Species == "Red duiker") %>%
  mutate(Censored = ifelse(Cens == 1, "Censored", "Observed")) %>%
  ggplot(aes(Stay, fill = Censored)) +
  geom_histogram(bins = 25, colour = "white", alpha = 0.9) +
  scale_fill_manual(values = c(Observed = "#4C72B0", Censored = "#DD8452")) +
  labs(x = "Staying time (s)", y = "Number of passes", fill = NULL) +
  theme_minimal(base_size = 12)
```

## Step 2: Passes and effort

For the original REST we need the total passes `Y` per station; for RAD-REST we
need the pass-count classes. `ct_rest_passes()` builds either with the `model`
argument.

```{r passes}
# Original REST: total passes per station
stations_rest <- ct_rest_passes(rest_detection, rest_station, model = "REST")

# RAD-REST: videos split into y_0, y_1, ... classes
stations_rad  <- ct_rest_passes(rest_detection, rest_station, model = "RAD-REST")

stations_rad %>% filter(Species == "Red duiker") %>% head(4)
```

The pass-count classes are what RAD-REST models. Most videos show a single pass,
a few show more, and the model learns this whole distribution rather than a
single hand-counted total.

```{r pass-plot, fig.height=3.5, fig.cap="Figure 2: Pass-count classes for Red duiker, summed over stations. RAD-REST models this distribution with a Dirichlet-multinomial to estimate the mean number of passes per video."}
stations_rad %>%
  filter(Species == "Red duiker") %>%
  summarise(across(starts_with("y_"), sum)) %>%
  tidyr::pivot_longer(everything(), names_to = "class", values_to = "videos") %>%
  mutate(passes = as.integer(sub("y_", "", class))) %>%
  ggplot(aes(passes, videos)) +
  geom_col(fill = "#4C72B0", width = 0.7) +
  labs(x = "Passes per video", y = "Number of videos") +
  theme_minimal(base_size = 12)
```

Effort (camera-active days) is approximated from the span of detections at each
station with `ct_rest_effort()`. Supply a real deployment table if you have one.

```{r effort}
stations_rest <- ct_rest_effort(rest_detection, stations_rest)
stations_rad  <- ct_rest_effort(rest_detection, stations_rad)

stations_rest %>% filter(Species == "Red duiker") %>%
  select(Station, Species, Y, Habitat, Effort) %>% head(4)
```

## Step 3: Activity

`ct_rest_activity()` keeps independent detections and converts their time of day
to radians, ready for the activity model.

```{r activity}
activity <- ct_rest_activity(rest_detection, independence_minutes = 30)
head(activity, 4)
```

Plotting the time of day reveals the species' diel pattern — here the clear
crepuscular (dawn/dusk) peaks that determine $p_{\text{act}}$.

```{r activity-plot, fig.height=3.5, fig.cap="Figure 3: Time-of-day of independent detections for the focal species. The activity proportion that enters the REST equation is estimated from this circular distribution, either by kernel density or a Bayesian von Mises mixture."}
activity %>%
  filter(Species == "Red duiker") %>%
  mutate(hour = time * 12 / pi) %>%
  ggplot(aes(hour)) +
  geom_density(fill = "#4C72B0", alpha = 0.5, adjust = 0.8) +
  scale_x_continuous(breaks = seq(0, 24, 6), limits = c(0, 24)) +
  labs(x = "Hour of day", y = "Activity density") +
  theme_minimal(base_size = 12)
```

## Step 4: Choosing the staying-time distribution {#choosing-the-staying-time-distribution}

Rather than guessing, compare candidate distributions by WAIC. `bayes_p` is a
posterior predictive check (values near 0.5 indicate good fit).

```{r select-stay, eval=FALSE}
ct_rest_select_stay(
  stay, species = "Red duiker",
  stay_distribution = c("lognormal", "gamma", "weibull", "exponential"),
  iterations = 3000, burnin = 1000, chains = 2, cores = 2
)
```

```text
── Staying-time model selection ──

# A tibble: 4 × 5
  model      family       random_effect  WAIC bayes_p
  <chr>      <chr>        <chr>          <dbl>   <dbl>
1 Stay ~ 1   lognormal    none            612.    0.49
2 Stay ~ 1   gamma        none            628.    0.41
3 Stay ~ 1   weibull      none            641.    0.38
4 Stay ~ 1   exponential  none            705.    0.12
```

Lognormal wins here, so we use it below.

## Step 5: Fit the REST model

`ct_fit_rest()` runs the whole joint model. Give it the three tables, the focal
species, the focal-area size in m², and the chosen distribution.

```{r fit-rest, eval=FALSE}
fit <- ct_fit_rest(
  stay_data = stay,
  station_data = stations_rest,
  activity_data = activity,
  species = "Red duiker",
  focal_area = 3.0, # focal-area size in m^2
  model = "REST",
  stay_distribution = "lognormal",
  iterations = 5000, burnin = 1000, chains = 3, cores = 3
)
fit
```

```text
── REST density estimation ──

# A tibble: 2 × 11
  Species    Station Variable   mean    sd lower median upper  Rhat n.eff     cv
  <chr>      <chr>   <chr>     <dbl> <dbl> <dbl>  <dbl> <dbl> <dbl> <int>  <dbl>
1 Red duiker All     density   29.5  6.73  18.7   28.6  46.1  1.00   485 0.229
2 Red duiker All     mean_stay  6.01 0.33   5.45   5.98  6.67  1.00   359 0.054
```

`density` is in **individuals per km²**, `mean_stay` in **seconds**, each with a
posterior mean, standard deviation, 95% credible interval, $\hat{R}$ and
effective sample size. Full posterior draws are in `fit$samples` (a
`coda::mcmc.list`) for trace plots with the `MCMCvis` package, and `fit$waic`
ranks any candidate density models.

## Step 6: Fit RAD-REST

Switching to RAD-REST needs only the pass-classified station table and
`model = "RAD-REST"`:

```{r fit-rad, eval=FALSE}
fit_rad <- ct_fit_rest(
  stay_data = stay,
  station_data = stations_rad,
  activity_data = activity,
  species = "Red duiker",
  focal_area = 3.0,
  model = "RAD-REST",
  stay_distribution = "lognormal"
)
fit_rad
```

```text
── RAD-REST density estimation ──

# A tibble: 3 × 11
  Species    Station Variable   mean    sd lower median upper  Rhat n.eff     cv
  <chr>      <chr>   <chr>     <dbl> <dbl> <dbl>  <dbl> <dbl> <dbl> <int>  <dbl>
1 Red duiker All     density   29.5  8.88  17.9   28.5  46.3  1.02   338 0.301
2 Red duiker All     mean_stay  5.99 0.34   5.38   5.98  6.73  1.00   296 0.057
3 Red duiker All     mean_pass  1.25 0.08   1.09   1.25  1.41  1.01   913 0.065
```

RAD-REST reports the extra `mean_pass` term (≈ 1.25 passes per video) and here
recovers essentially the same density as REST — reassuring, since the two
differ only in how the passes are counted.

## Going further: covariates, random effects, and activity uncertainty

`ct_fit_rest()` accepts model formulas and a random-effect term, so density and
staying time can vary with habitat or other covariates, and the von Mises
mixture can replace the kernel when you want activity uncertainty propagated:

```{r extras, eval=FALSE}
# Density varies by habitat; station-level random effect on staying time;
# Bayesian activity estimation.
ct_fit_rest(
  stay_data = stay,
  station_data = stations_rest,
  activity_data = activity,
  species = "Red duiker",
  focal_area = "FocalArea", # a per-camera focal-area column in station_data
  model = "REST",
  density_formula = ~ Habitat, # density covariate
  stay_random_effect = "Station", # random effect on staying time
  activity_method = "mixture" # propagate activity uncertainty
)
```

With a covariate in `density_formula`, the summary returns a per-station density
column instead of a single `All` row; `focal_area` may be a single number or the
name of a per-camera column in `station_data`.

# Interpreting the estimates

- **Density units.** `density` is individuals per km². Multiply by your study
  area (in km²) for abundance.
- **Convergence.** Check that $\hat{R} \approx 1.00$ and `n.eff` is comfortably
  in the hundreds; if not, raise `iterations`/`chains`. The examples above use
  short runs for speed.
- **Model choice.** Use `ct_rest_select_stay()` to pick the staying-time
  distribution, and `compare_models = TRUE` in `ct_fit_rest()` to rank density
  covariate sets by WAIC (`fit$waic`).
- **REST vs RAD-REST.** Prefer RAD-REST when counting every pass is impractical
  or error-prone; prefer REST when you already have reliable total pass counts.
  They should agree when both are well-specified.

# References

Nakashima, Y., Fukasawa, K., & Samejima, H. (2018). Estimating animal density
without individual recognition using information derived from camera traps.
*Journal of Applied Ecology*, 55(2), 735–744.
<https://doi.org/10.1111/1365-2664.13059>

Nakashima, Y., Yajima, G., & Matsuoka, R. (2026). Reducing data processing effort in camera trap density estimation: Extending the REST model by explicitly modelling animal detection processes. *Methods in Ecology and Evolution*, 2026(December 2025), 850–862. https://doi.org/10.1111/2041-210x.70248

Rowcliffe, J. M., Kays, R., Kranstauber, B., Carbone, C., & Jansen, P. A.
(2014). Quantifying levels of animal activity using camera trap data. *Methods
in Ecology and Evolution*, 5(11), 1170–1179.
<https://doi.org/10.1111/2041-210X.12278>
